---
title: "SWNE analysis of hematopoiesis chromatin accessibility"
author: "Yan Wu"
date: "4/26/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a walkthrough on how to use SWNE to visualize a hematopoiesis chromatin accessibility trajectory from [Buenrostro et al, 2018](https://doi.org/10.1016/j.cell.2018.03.074), starting from a [cisTopic](https://github.com/aertslab/cisTopic) object that has already been processed using the cisTopic pipeline. That object can be downloaded [here](ftp://genome-miner.ucsd.edu/swne_files/scATAC_hemato_cisTopic.Robj).

Before we start, you'll need some additional packages for TF binding site annotation and analysis.

```{r eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("motifmatchr", version = "3.8")
BiocManager::install("BSgenome.Hsapiens.UCSC.hg19", version = "3.8")
BiocManager::install("TxDb.Hsapiens.UCSC.hg19.knownGene", version = "3.8")
BiocManager::install("org.Hs.eg.db", version = "3.8")
BiocManager::install("SummarizedExperiment", version = "3.8")
BiocManager::install("TFBSTools", version = "3.8")
BiocManager::install("chromVAR")

if(!require(devtools)){ 
  install.packages("devtools") # If not already installed; 
}
devtools::install_github("GreenleafLab/chromVARmotifs")
```

Load the required libraries and data. The data has already been preprocessed with  using their standard pipeline. 

```{r message=FALSE,warning=FALSE}
library(cisTopic)
library(swne)
library(chromVAR)
library(chromVARmotifs)
library(motifmatchr)
library(SummarizedExperiment)
library(org.Hs.eg.db)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
library(BSgenome.Hsapiens.UCSC.hg19)
library(TFBSTools)

load("~/swne/Data/scATAC_hemato_cisTopic.RData")
```


We can run SWNE directly on the cisTopicObject. Increasing `snn.k` and decreasing `snn.exp` will pull more cells onto the primary axes of variation (in this case the monocyte/lymphoid and erythrocyte trajectories) at the expense of some cell type separation.
```{r message=FALSE,warning=FALSE}
## Run SWNE
swne.emb <- RunSWNE(cisTopicObject, alpha.ex = 1.25, snn.exp = 1, 
                    n_pull = 3, snn.k = 50, prune.SNN = 1/20)
swne.emb$H.coords$name <- ""
```


We need to annotate the peaks in order to embed promoter/binding site accessibility onto the visualization
```{r message=FALSE,warning=FALSE}
## Annotate topics
cisTopicObject <- getRegionsScores(cisTopicObject)
cisTopicObject <- annotateRegions(cisTopicObject, txdb = TxDb.Hsapiens.UCSC.hg19.knownGene,
                                  annoDb = "org.Hs.eg.db")
```


Next we'll embed the promoter accessibility of some key TFs
```{r message=FALSE,warning=FALSE}
## Embed promoter accessibility of key TFs
tfs.embed <- c("GATA1", "CEBPB")
swne.emb <- EmbedPromoters(swne.emb, cisTopicObject, tfs.embed, n_pull = 3, promoterOnly = T)
```


We also want to look at the accessibility of the TF binding sites. This requires us to annotate which peaks overlap with which TF's binding sites.
```{r message=FALSE,warning=FALSE}
peak.gr <- cisTopicObject@region.ranges
fragment_counts <- SummarizedExperiment(assays = list(counts = cisTopicObject@binary.count.matrix),
                                        rowRanges = peak.gr)
fragment_counts <- addGCBias(fragment_counts, genome = BSgenome.Hsapiens.UCSC.hg19)

data("human_pwms_v2") ## Get position weight matrices from chromVARMotifs
motif_ix <- matchMotifs(human_pwms_v2, fragment_counts, genome = BSgenome.Hsapiens.UCSC.hg19)
motif_ix_mat <- assay(motif_ix)*1

## Extract the TF symbol from the motif names
colnames(motif_ix_mat) <- sapply(colnames(motif_ix_mat), ExtractField, field = 3, delim = "_")
```


Now we can embed the binding site accessibility of CEBPB and GATA1 and generate our final SWNE plot.
```{r message=FALSE,warning=FALSE}
swne.emb <- EmbedTFBS(swne.emb, cisTopicObject, motif_ix_mat, tfs.embed, n_pull = 3,
                      overwrite = F)

plot.seed <- 512223535
PlotSWNE(swne.emb, sample.groups = clusters, pt.size = 1.25, alpha = 0.5, 
         do.label = T, seed = plot.seed)
```
